External priors for the next generation of CMB experiments.

### 2015-12-07 Notes/N_eff_prior_paper_v3.tex
small edits

\date{\today}
\begin{abstract}
Planned cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, and dark energy. 
Indeed the low level of noise, together with improved angular resolution, will increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure. Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments. Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment on neutrino and dark energy properties.
We find that CMB S4 constraints on neutrino mass could be strongly enhanced by external constraints on the cold dark matter density $\Omega_{c}h^{2}$ and the Hubble constant $H_{0}$. If the largest scales ($\ell<50$) will not be measured, an external prior on the primordial amplitude $A_{s}$ or the optical depth $\tau$ will also be important. Furthermore a CMB constraint on the number of relativistic degrees of freedom, $N_{\rm eff}$, will benefit from an external prior on the spectral index $n_{s}$ and the baryon energy density $\Omega_{b}h^{2}$. Finally, an external prior on $H_{0}$ will help constrain the dark energy equation of state ($w$).
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
Since their earliest incarnations, Cosmic Microwave Background (CMB) experiments have been crucial in furthering our understanding of the universe. They will maintain their role in the future, thanks to the unprecedented low level of noise and high resolution expected in the planned Stage IV (S4) experiment~\cite{2013arXiv1309.5383A}.
The recent past is reassuring. For example, every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P, planck-collaboration:2014, 2003ApJS..148..175S, 2000ApJ...545L...5H,2000Natur.404..955D}. 
These led us from the first detection of anisotropy in the CMB temperature with COBE to a cosmic variance limited measurement of several acoustic peaks with Planck. With the improved sensitivity we have extended our understanding of the universe so that we now have solid evidence for a flat geometry and a well tested model $\Lambda$CDM with percent level constraints on its parameters.
The same success has characterized ground-based CMB experiments, where, e.g., the first detection of polarization anisotropy by DASI \cite{2002ApJ...568...38H} has been followed by a series of precise measurements of the damping tail and maps of the projected gravitational potential by, e.g., SPT~\cite{2011ApJ...743...28K} and ACT~\cite{2011ApJ...739...52D}.
The planned CMB-S4 experiment will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction. As has happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, dark energy, and neutrinos.

@@

The CMB is sensitive also to the properties of dark energy (see, e.g., \cite{2010MNRAS.405.2639J}). Thanks to the current generation of experiments we know with extraordinary accuracy its energy density. The challenge for the next generation is to reveal the nature of this mysterious component. For example, a crucial step to identifying the mechanism driving cosmic acceleration will be to investigate any possible deviations in the equation of state, the ratio of pressure and energy density, from the value $w=-1$ predicted by a cosmological constant. 
Dark energy affects the CMB because it alters the universe's expansion and it consequently changes the distance to the last scattering surface. Furthermore different dark energy models lead to different rates of growth of large scale structure which are tested by CMB lensing. 
However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$. As with the neutrino sector, dark energy constraints from the CMB 
will be improved by external experiments.

In the case of both neutrinos and dark energy then, an important lingering question is: What external information can be used to improve projected constraints by breaking degeneracies?
Here we study the dependance of projected constraints from CMB on the external priors assumed. For all the parameters of the standard $\Lambda$CDM with massive neutrinos ($\Omega_ch^2, \Omega_bh^2, A_s, n_s, \tau, H_0, \sum m_\nu$) together with extensions ($N_{\rm eff}$ and $w$) to the neutrino and dark energy sectors, we quantify the CMB constraints as a function of external priors. This extends the work of \citet{wu:2014}, which worked with a few fixed external priors, by quantifying the extent to which external information will improve the constraining power of a CMB-S4 experiment.

This paper is organized as follow: in \refsec{methods} we introduce the technique and assumptions we use to derive the effect of external priors on the CMB parameter constraints. In \refsec{results} we will describe our results and we then conclude with a discussion of them in \refsec{conclusions}.

@@

We also include information from current Baryon Acoustic Oscillation experiments and projected constraints from low-$l$ Planck polarization. 
As current BAO data (labeled BAO15), we follow \cite{allison:2015} and we take into account 6dFGRS \cite{beutler:2011}, SDSS MGS \cite{ross:2015} together with LOW-Z and C-MASS \cite{anderson:2014}.
The noise assumed for Planck corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.

@@

\reffig{prior_omeganuh2} shows the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors. 
Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 25 meV, corresponding to a 2-sigma detection even in the worst case scenario. Adding in external priors helps this significantly though, as discussed, for example, in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}. 
There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI. This is represented by the blue band in \reffig{prior_omeganuh2}. In addition to showing the implications of introducing this fixed external constraint, \reffig{prior_omeganuh2} generalizes the extent to which external information will improve the neutrino mass constraints. For example, the red curve shows the improvement as additional information on the Hubble constant is applied. The projected, marginalized constraint on $H_0$ from just CMB-S4+Planck+BAO15 is $0.5\%$ (and similarly the values for the other parameters are given in the inset). So imposing {\it external} information with this value of $\sigma_{H_0}/H_0=0.005$ corresponds to 1 on the $x$-axis. This would improve the neutrino mass constraint to 17 meV. Similarly, introducing an external constraint on the dark matter density equal to its internal constraint of 0.3\% (so again 1 on the $x$-axis) improves the projected neutrino mass constraint to 14 meV. It is not surprising that this is equal to the constraint obtained with DESI because the BAO distances in this model depends most sensitively on $\Omega_ch^2$ and $H_{0}$.

The physical reason underlying this improvement is simple: lensing constrains $f_\nu$, the ratio of the neutrino density to the matter density. In order to extract the neutrino density (and therefore the neutrino mass), we need to know the matter density. A prior on $H_{0}$ helps for a related, but subtle reason. 
The position of the CMB peaks, well constrained by measurement, in a flat universe is a function $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} 
. For this reason an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors. 
The left-hand side of \reffig{prior_omeganuh2} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_snow_mass_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
\caption{Same as \reffig{prior_omeganuh2} but with $\pmb{\ell_{\rm min}>50}$. Here $A_{s}$ and $\tau$ priors will also helps because of their degeneracy with $\sum m_{\nu}$.}
\label{fig:prior_omeganuh2_tau}
\end{center}
\end{figure}

@@

Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and, as a consequence, between $\sum m_\nu$ and $\tau$. This can be understood as follows.
The CMB lensing reconstruction is noisy at scales larger than the neutrino free streaming scale (see Fig. 5 of \cite{2013arXiv1309.5383A}). For this reason it is hard to measure the unsuppressed CMB lensing power spectra that is then compared to the small scale suppressed one to constrain the values of $f_{\nu}$. The unsuppressed amplitude of the lensing spectrum is actually better constrained by the primordial amplitude $A_{s}$, which is probed by the primordial CMB. Unfortunately, the CMB spectra are sensitive to $A_{s}\exp^{-2\tau}$, so 
a measurement of the optical depth is required to infer $A_{s}$ and therefore to tighten the constraints on $\sum m_\nu$. If CMB S4 will not be designed to measure scales $\ell<50$ an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like PIXIE \cite{kogut:2011} or balloon like SPIDER \cite{filippini:2010}, will be needed to fully exploit the CMB constraining power.

@@

The sensitivity to relativistic degrees of freedom comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also strongly affect this part of the spectrum, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are the most relevant to improce $N_{\rm eff}$ constraint. As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to ensure a 3-sigma detection of the partial decoupling prediction. The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly.

@@

\reffig{prior_w} shows that without any priors, CMB experiments will do better than current constraints, which hover around 10\%. However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ by a factor of 2. 
This is primarily due to the fact that CMB constrains $w$ through a precise measurement of the comoving distance to the last scattering surface. However that can be kept fixed while varying $w$ by accordingly changing the value of $H_{0}$ (and $\Omega_{\Lambda}$ to keep the universe flat). Indeed in a two dimensional ($H_{0},w$) space the CMB constraint approximately lies on the region of constant large scattering surface distance as shown in \reffig{ellipse_w_H0}.
Notice that an external prior on $H_{0}$ two times more accurate than CMB S4 can improve the CMB constraint almost by a factor of 2 from a $4\%$ error to a $2\%$ level.

>>>>

\date{\today}
\begin{abstract}
Planned cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, and dark energy. 
The low level of noise, together with improved angular resolution, will increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure. Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments. Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage IV (S4) experiment on neutrino and dark energy properties.
We find that CMB S4 constraints on neutrino mass could be strongly enhanced by external constraints on the cold dark matter density $\Omega_{c}h^{2}$ and the Hubble constant $H_{0}$. If polarization on the largest scales ($\ell<50$) will not be measured, an external prior on the primordial amplitude $A_{s}$ or the optical depth $\tau$ will also be important. A CMB constraint on the number of relativistic degrees of freedom, $N_{\rm eff}$, will benefit from an external prior on the spectral index $n_{s}$ and the baryon energy density $\Omega_{b}h^{2}$. Finally, an external prior on $H_{0}$ will help constrain the dark energy equation of state ($w$).
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
Since their earliest incarnations, Cosmic Microwave Background (CMB) experiments have been crucial in furthering our understanding of the universe. They will maintain their role in the future, thanks to the unprecedented low level of noise and high resolution expected in the planned Stage IV (S4) experiment~\cite{2013arXiv1309.5383A}.
The recent past is reassuring. For example, every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P, planck-collaboration:2014, 2003ApJS..148..175S, 2000ApJ...545L...5H,2000Natur.404..955D}. 
These led us from the first detection of anisotropy in the CMB temperature with COBE to a cosmic variance limited measurement of several acoustic peaks with Planck. With the improved sensitivity we have extended our understanding of the universe so that we now have solid evidence for a flat geometry and a well tested model $\Lambda$CDM with percent level constraints on its parameters.
The same success has characterized ground-based CMB experiments, where, e.g., the first detection of polarization anisotropy by DASI \cite{2002ApJ...568...38H} has been followed by a series of precise measurements of the damping tail and maps of the projected gravitational potential by, e.g., SPT~\cite{2011ApJ...743...28K} and ACT~\cite{2011ApJ...739...52D}.
The planned CMB-S4 experiment will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction. As has happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of dark matter, inflation, dark energy, neutrinos, and other  Beyond the Standard Model physics.

@@

The CMB is sensitive also to the properties of dark energy (see, e.g., \cite{2010MNRAS.405.2639J}). Thanks to the current generation of experiments we know with extraordinary accuracy its energy density. The challenge for the next generation is to reveal the nature of this mysterious component. For example, a crucial step to identifying the mechanism driving cosmic acceleration will be to investigate any possible deviations in the equation of state, the ratio of pressure and energy density, from the value $w=-1$ predicted by a cosmological constant. 
Dark energy affects the CMB because it alters the universe's expansion and it consequently changes the distance to the last scattering surface. Furthermore different dark energy models lead to different rates of growth of large scale structure which are tested by CMB lensing. 
However, dark energy properties are strongly degenerate with other geometrical parameters such as $H_{0}$. As with the neutrino sector, dark energy constraints from the CMB 
will be improved by external experiments.

In the cases of both neutrinos and dark energy then, an important lingering question is: What external information can be used to improve projected constraints by breaking degeneracies?
Here we study the dependance of projected constraints from CMB on the external priors assumed. For all the parameters of the standard $\Lambda$CDM with massive neutrinos ($\Omega_ch^2, \Omega_bh^2, A_s, n_s, \tau, H_0, \sum m_\nu$) together with extensions ($N_{\rm eff}$ and $w$) to the neutrino and dark energy sectors, we quantify the CMB constraints as a function of external priors. This extends the work of \citet{wu:2014}, which worked with a few fixed external priors, by quantifying the extent to which external information will improve the constraining power of a CMB-S4 experiment.

This paper is organized as follow: in \refsec{methods} we introduce the technique and assumptions used to derive the effect of external priors on the CMB parameter constraints. In \refsec{results} we will describe our results and we then conclude with a discussion of them in \refsec{conclusions}.

@@

We also include information from current Baryon Acoustic Oscillation experiments and projected constraints from low-$l$ Planck polarization. 
For current BAO data (labeled BAO15), we follow \cite{allison:2015} and include 6dFGRS \cite{beutler:2011}, SDSS MGS \cite{ross:2015} together with LOW-Z and C-MASS \cite{anderson:2014}.
The noise assumed for Planck corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.

@@

\reffig{prior_omeganuh2} shows the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors. 
Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 25 meV, corresponding to a 2-sigma detection even in the worst case scenario. Adding in external priors helps this significantly though, as discussed, for example, in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}. 
There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI. This is represented by the blue band in \reffig{prior_omeganuh2}. In addition to showing the implications of introducing this fixed external constraint, \reffig{prior_omeganuh2} generalizes the extent to which external information will improve the neutrino mass constraints. For example, the red curve shows the improvement as additional information on the Hubble constant is applied. The projected, marginalized constraint on $H_0$ from just CMB-S4+Planck+BAO15 is $0.6\%$ (and similarly the values for the other parameters are given below the figure). So imposing {\it external} information with this value of $\sigma_{H_0}/H_0=0.006$ corresponds to 1 on the $x$-axis. This would improve the neutrino mass constraint to 17 meV. Similarly, introducing an external constraint on the dark matter density equal to its internal constraint of 0.4\% (so again 1 on the $x$-axis) improves the projected neutrino mass constraint to 20 meV. It is not surprising that this is close to the constraint obtained with DESI because the BAO distances in this model depends most sensitively on $\Omega_ch^2$ and $H_{0}$.

The physical reason underlying this improvement is simple: lensing constrains $f_\nu$, the ratio of the neutrino density to the matter density. In order to extract the neutrino density (and therefore the neutrino mass), we need to know the matter density. A prior on $H_{0}$ helps for a related, but subtle reason. 
The position of the CMB peaks, well constrained by measurement, in a flat universe depends on the combination $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} 
. Therefore, an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors. 
The left-hand side of \reffig{prior_omeganuh2} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_snow_mass_lmin=50_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
\caption{Same as \reffig{prior_omeganuh2} but with $\pmb{\ell_{\rm min}>50}$. Here $A_{s}$ and $\tau$ priors also help because of their degeneracy with $\sum m_{\nu}$.}
\label{fig:prior_omeganuh2_tau}
\end{center}
\end{figure}

@@

Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and, as a consequence, between $\sum m_\nu$ and $\tau$. This can be understood as follows.
The CMB lensing reconstruction is noisy at scales larger than the neutrino free streaming scale (see Fig. 5 of \cite{2013arXiv1309.5383A}). For this reason it is hard to measure the unsuppressed CMB lensing power spectra that is then compared to the small scale suppressed one to constrain the values of $f_{\nu}$. The unsuppressed amplitude of the lensing spectrum is actually better constrained by the primordial amplitude $A_{s}$, which is probed by the primordial CMB. Unfortunately, the CMB spectra are sensitive to $A_{s}\exp^{-2\tau}$, so 
a measurement of the optical depth is required to infer $A_{s}$ and therefore to tighten the constraints on $\sum m_\nu$. If CMB S4 will not be designed to measure scales $\ell<50$ an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like PIXIE \cite{kogut:2011} or balloon-borne experiments like SPIDER \cite{filippini:2010}, will be needed to fully exploit the CMB constraining power.

@@

The sensitivity to relativistic degrees of freedom comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also strongly affect this part of the spectrum, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are the most relevant to improve the $N_{\rm eff}$ constraint. As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to get close to a 4-sigma detection of the partial decoupling prediction. The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly.

@@

\reffig{prior_w} shows that without any priors, CMB experiments will do better than current constraints, which hover around 10\%. However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ by a factor of 2. 
This is primarily due to the fact that CMB constrains $w$ through a precise measurement of the comoving distance to the last scattering surface. However that can be kept fixed while varying $w$ by accordingly changing the value of $H_{0}$ (and $\Omega_{\Lambda}$ to keep the universe flat). Indeed in a two dimensional ($H_{0},w$) space the CMB constraint approximately lies on the region of constant large scattering surface distance as shown in \reffig{ellipse_w_H0}.
Notice that an external prior on $H_{0}$ two times more accurate than CMB S4 can improve the CMB constraint on $w$ by a factor of two.
